Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-11-13 papers/histogram/sad.tex
try to figure out what Poulaine et al. actually did, requires a script
clarify Poulaine paragraph
major edits on LJ31 bits, totally new figure

\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\jpsays}[1]{{\color{red} [\blue{Jordan:} \emph{#1}]}}
\newcommandx{\jpcom}[2][1=inline]{\todo[linecolor=gray,backgroundcolor=light-gray,bordercolor=dark-gray,#1]{\textbf{Jordan says:} #2} }
\begin{document}

@@

\begin{abstract}
  \jpsays{edit to include LJ clusters}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interest, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

\jpcom{edit some here!}
As a test case, we consider the square-well fluid i.e. a system of particles
whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The square-well
potential is an ideal test-bed as it is a simple model for a liquid, which
includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.  The potential
$U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

\jpcom{Describe Lennard-Jones clusters we test SAD ability to handle complex
energy landscapes. We choose 31 and 38 due to complex funnels In the particular
case of n=38 there exist two funnels at low energies representing octahedral (Oh)
and icosahedral (ico) energy minima [5] leading to a pronounced size-specific
behavior at low temperatures, which is especially difficult to simulate. The
first numerical success}

In this work, we compare four flat histogram methods.  We outline the
general workings of each algorithm that we developed in detail while
summarizing algorithms that were developed in other works.  The
following methods are discussed and simulated for the square-well
fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte Carlo (SAMC), and SAD.
\jpcom{In addition, we test the ability of SAD to accurately and efficiently
samples low-temperature solid-solid transitions for Lennard-Jones clusters.}

@@

@@

\subsection{A periodic system with 50 atoms}

For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from $-248$ to $-120$.  The number of important energy states
for this system is therefore $N_S = 129$.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

@@

\subsection{A periodic system with 256 atoms}
Next we will introduce a considerably larger simulation containing 256
atoms which has a maximum entropy about 1500 greater than its minimum.
This makes exploring the entire range of energies extremely expensive,
and strongly favors the methods that restrict the energy (or
temperature) range of interest.  For this simulation, we chose a much
higher minimum reduced temperature of $1.0$, which corresponds to an
interesting energy range from $-915$ to $-509$.  The number of
important energy states for this system is then $N_S = 407$.  The
minimum entropy over this energy range is just 395 less than the
maximum, corresponding to a ratio of only $10^{118}$ between the
highest and lowest density of states.

@@

@@

\subsection{Lennard-Jones clusters with 31 atoms}

Clusters of Lennard-Jones atoms have been frequently used for testing
Monte Carlo algorithms~\davidsays{add a few citations}.  The 31-atom
cluster is popular because it features a two-funnel energy landscape
with a significant barrier between the two low-energy
states~\davidsays{add citations}.

Poulain \emph{et al.} found very poor convergence for a variety of
modified WL methods on the LJ31
cluster~\cite{poulain2006performances}.  Sadly, they don't mention the
energy range or bin size that was used in their simulation, so their
results cannot be reproduced.  They do mention that they use 1000
energy bins, which if they used an energy range of
$[-133.58\epsilon,0]$ would correspond to a bin size of $\Delta
E\approx 0.13\epsilon$, which we find to be too large to accurately
predict the heat capacity peak at $0.03\epsilon/k_B$.

@@

Dependence on bin width.  Figure~\ref{fig:lj-cv-error} shows that the
heat capacity converges just as well for bin widths varying over two
orders of magnitude.  WL and $1/t$-WL show similar behavior, so this
is not new, but is encouraging.

@@

>>>>

\newcommand{\davidsays}[1]{{\color{blue} [\green{David:} \emph{#1}]}}
\newcommand{\jpsays}[1]{{\color{red} [\blue{Jordan:} \emph{#1}]}}
\newcommandx{\jpcom}[2][1=inline]{\todo[linecolor=gray,backgroundcolor=light-gray,bordercolor=dark-gray,#1]{\textbf{Jordan says:} #2} }
\begin{document}

@@

\begin{abstract}
  \jpsays{edit to include LJ clusters}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  the 31-atom Lennard-Jones cluster and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interest, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more powerful in the common case in which the range
  of energies is not known in advance.
\end{abstract}

@@

In this work, we compare four flat histogram methods.  We outline the
general workings of each algorithm that we developed in detail while
summarizing algorithms that were developed in other works.  The
following methods are discussed and simulated for the square-well
fluid and the 31-atom Lennard-Jones cluster: Wang-Landau (WL),
$1/t$-Wang-Landau ($1/t$-WL), Stochastic Approximation Monte Carlo
(SAMC), and SAD.

@@

\subsection{Square-well fluid}
\davidsays{Edit this paragraph: introduce novelty of square well fluid
in particular, and importance.}
As our first test case, we consider the square-well fluid i.e. a
system of particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.  The
potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

@@

\subsubsection{50 square-well atoms}
\davidsays{check this periodic}
For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from $-248$ to $-120$.  The number of important energy states
for this system is therefore $N_S = 129$.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

@@

\subsubsection{256 square-well atoms}
\davidsays{check this periodic}
Next we will introduce a considerably larger simulation containing 256
atoms which has a maximum entropy about 1500 greater than its minimum.
This makes exploring the entire range of energies extremely expensive,
and strongly favors the methods that restrict the energy (or
temperature) range of interest.  For this simulation, we chose a much
higher minimum reduced temperature of $1.0$, which corresponds to an
interesting energy range from $-915$ to $-509$.  The number of
important energy states for this system is then $N_S = 407$.  The
minimum entropy over this energy range is just 395 less than the
maximum, corresponding to a ratio of only $10^{118}$ between the
highest and lowest density of states.

@@

\subsection{31-atom Lennard-Jones cluster}

In order to explore both the bin-size behavior and low-temperature
effectiveness of our algorithm, we chose to study the 31-atom cluster
(LJ31).
Clusters of Lennard-Jones atoms have been frequently used for testing
Monte Carlo algorithms~\davidsays{add a few citations}.  The LJ31
cluster is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\davidsays{add citations}.
In particular, the LJ31 cluster features a solid-solid phase
transition at a temperature of around $T=0.03\epsilon/k_B$ that is
challenging to converge~\davidsays{cite}.  Poulain \emph{et al.}
tested a variety of variants of WL on LJ31 and found
\davidsays{summarize}~\cite{poulain2006performances}.

\jpcom{This \P\ should be commented once summarized.
Poulain \emph{et al.} found very poor convergence for a variety of
modified WL methods on the LJ31
cluster~\cite{poulain2006performances}.  Sadly, they don't mention the
energy range or bin size that was used in their simulation, so their
results cannot be reproduced.  They do mention that they use 1000
energy bins, which if they used an energy range of
$[-133.58\epsilon,0]$ would correspond to a bin size of $\Delta
E\approx 0.13\epsilon$, which we find to be too large to accurately
predict the heat capacity peak at $0.03\epsilon/k_B$.  Looks like they
only ran $\sim 10^{10}$ moves. We can't converge in that few moves either.}

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster as
a function of temperature.  There are two peaks, a large peak at 0.2?
corresponding to melting temperature, and a small peak around 0.03 for
the solid-solid transition from a mackay to antimackay
wahatever~\davidsays{cite}.  We keep our atoms in a spherical box with
radius 2.5$\sigma$, as others do~\davidsays{cite}.  We find that using
a larger box (with radius $5\sigma$) has a large effect on the melting
peak, and a smaller but still significant effect on the solid-solid
transition.  Since our focus is on the convergence rate, we focused on
the smaller box size.  We further restricted our simulations to
negative energies, as the microcanonical energy at infinite
temperature is rather high for such a small box.

@@

@@

Dependence on bin width.  Figure~\ref{fig:lj-cv-error} shows that the
heat capacity converges just as well for bin widths varying over two
orders of magnitude, although for bins as large as $\Delta
E=0.1\epsilon$ the error in the heat capacity is significant, as is
apparent in Fig~\ref{fig:lj-cv}.  The entropy itself converges much
more slowly for a smaller $\Delta E$, but when summing over energies
the statistical fluctuations in the entropy cancel. WL and $1/t$-WL
show similar behavior, so this is not new, but is encouraging.

@@

SAMC is a major pain.  The $t_0$ value needs to be adjusted when
$\Delta E$ is changed or when the energy range is changed.  The energy
range has the same downsides as the WL methods, although it manifests
differently, with failure to converge as the problem, rather than
failure for $\gamma_t$ to decrease.
